---
title: "Example script to highlight the usage of the `cytomapper` package"
author: "Nils Eling"
date: "`r Sys.Date()`"
output:     
    md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script serves as an example to highlight the usage of the [cytomapper](https://www.bioconductor.org/packages/release/bioc/html/cytomapper.html) R/Bioconductor package.
It was written for the [Indiana O’Brien Center Microscopy Workshop](http://static.medicine.iupui.edu/obrien/2021Schedule.pdf)
in June 2021.

## Prerequisites

To use this script, first you will need to [install R](https://www.r-project.org/) and install the `cytomapper` package via:

```{r install-cytomapper, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("cytomapper")
```

For convenience, please open the provided `cytomapper_demos.Rproj` file.
This will correctly set the working directory.

## Reading in example data

In the first instance, we will read in the example data provided in this repository.
To read in images, the `cytomapper` package provides the `loadImages` function.

```{r read-in-data, message=FALSE}
library(cytomapper)

# Read in multi-channel images
images <- loadImages("../data/images/", pattern = ".tiff")

# Read in masks
masks <- loadImages("../data/masks/", pattern = ".tiff", as.is = TRUE)

# Read in single-cell data
sce <- readRDS("../data/sce.rds")
```

## Add needed metadata

```{r}
# Add channel names
channelNames(images) <- rownames(sce)

# Add image name to metadata
mcols(images) <- mcols(masks) <- DataFrame(ImageName = c("E30", "G23", "J01"))
```

## Visualize multi-channel images

```{r}
plotPixels(image = images,
           colour_by = c("PIN", "CD4", "CD8a"), 
           colour = list(PIN = c("black", "yellow"),
                         CD4 = c("black", "blue"),
                         CD8a = c("black", "red")),
           bcg = list(PIN = c(0, 10, 1),
                      CD4 = c(0, 8, 1),
                      CD8a = c(0, 10, 1)),
           image_title = list(text = c("Non-diabetic",
                                       "Recent onset T1D",
                                       "Long duration T1D")),
           scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m")))
```

## Visualize segmented cells

```{r}
cur_sce <- sce[,sce$CellType %in% c("beta", "alpha", "delta", "Tc", "Th")]

plotCells(mask = masks,
          object = cur_sce,
          cell_id = "CellNumber",
          img_id = "ImageName",
          colour_by = "CellType",
          image_title = list(text = c("Non-diabetic",
                                      "Recent onset T1D",
                                      "Long duration T1D"),
                             colour = "black"),
          scale_bar = list(length = 100,
                           label = expression("100 " ~ mu * "m"),
                           colour = "black"),
          missing_colour = "white",
          background_colour = "gray")
```

## Outline cells on images

```{r}
plotPixels(image = images,
           object = cur_sce,
           mask = masks,
           cell_id = "CellNumber",
           img_id = "ImageName",
           colour_by = c("PIN", "CD4", "CD8a"), 
           outline_by = "CellType",
           colour = list(PIN = c("black", "yellow"),
                         CD4 = c("black", "blue"),
                         CD8a = c("black", "red")),
           bcg = list(PIN = c(0, 10, 1),
                      CD4 = c(0, 8, 1),
                      CD8a = c(0, 10, 1)),
           image_title = list(text = c("Non-diabetic",
                                       "Recent onset T1D",
                                       "Long duration T1D")),
           scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m")),
           thick = TRUE)
```

## Gate cells on images

```{r}
if (interactive()) {
    cytomapperShiny(sce, mask = masks, image = images, 
                    cell_id = "CellNumber", img_id = "ImageName")
}
```

## Session info

```{r sessionInfo}
sessionInfo()
```