---
title: "Bioc2021: Visualisation of highly-multiplexed imaging data in R"
date: "`r BiocStyle::doc_date()`"
author:
- name: Nils Eling 
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: nils.eling@dqbm.uzh.ch
- name: Nicolas Damond
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
- name: Tobias Hoch
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
- name: Bernd Bodenmiller
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
output:
    BiocStyle::html_document:
        toc_float: yes
        pandoc_args: [
            "--output=index.html"
            ]
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(dirname(inputFile),'/index.html')) })
abstract: |
    Highly multiplexed imaging cytometry acquires the single-cell expression of
    selected proteins in a spatially-resolved fashion. These measurements can be
    visualized across multiple length-scales. First, pixel-level intensities
    represent the spatial distributions of feature expression with highest
    resolution. Second, after segmentation, expression values or cell-level
    metadata (e.g. cell-type information) can be visualized on segmented cell
    areas. This package contains functions for the visualization of multiplexed
    read-outs and cell-level information obtained by multiplexed imaging
    cytometry. The main functions of this package allow 1. the visualization of
    pixel-level information across multiple channels and 2. the display of
    cell-level information (expression and/or metadata) on segmentation masks.
vignette: |
    %\VignetteIndexEntry{"Visualization of imaging cytometry data in R"}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

`r icons::fontawesome('twitter')` [@NilsEling](https://twitter.com/NilsEling)  
`r icons::fontawesome('github')` [@nilseling](https://github.com/nilseling)

# Data and code availability

To follow this tutorial, please visit
[https://github.com/BodenmillerGroup/cytomapper_demos/docs](https://github.com/BodenmillerGroup/cytomapper_demos/docs).
The compiled .html of this workshop is hosted at:
[https://bodenmillergroup.github.io/cytomapper_demos](https://bodenmillergroup.github.io/cytomapper_demos).
The
[cytomapper](https://www.bioconductor.org/packages/release/bioc/html/cytomapper.html)
package can be installed via:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("cytomapper")
```

To reproduce the analysis, clone the repository:

```
git clone https://github.com/BodenmillerGroup/cytomapper_demos.git
```

and open the `Bioc2021_workshop.Rmd` file in the `docs` folder.

We provide three images, their segmentation masks and the quantified single-cell
data in form of a `SingleCellExperiment` object in the `data` folder. The data
is taken from [A map of human type 1 diabetes progression by imaging mass
cytometry](https://www.sciencedirect.com/science/article/pii/S1550413118306910).

# Introduction

The analysis and visualization of highly-multiplexed imaging data relies on three
types of data structures:

* Multi-channel images realized as three dimensional arrays where the numeric
entry of each voxel represents the intensity of the pixel at position `x` and
`y` for channel `c`.

* Single-channel segmentation masks realized as matrices where sets of pixels
with the same integer ID represent individual objects (here, these are segmented
cells)

* A table containing the quantified features per cell and channel (e.g. mean
pixel intensity)

The `cytomapper` package handles these data types using objects of the following
S4 classes:

* Multiple multi-channel images are stored in form of `EBImage::Image` objects
within a `cytomapper::CytoImageList` object

* Multiple single-channel segmentation masks are stored in form of
`EBImage::Image` objects within a `cytomapper::CytoImageList` object

* Per-cell intensity measures and cell-/channel-specific metadata are stored in
a `SingleCellExperiment::SingleCellExperiment` container

![cytomapper overview figure. A) The plotCells function combines a SingleCellExperiment and CytoImageList object to visualize marker expression or cell-specific metadata on segmentation masks. B) The plotPixels function requires a CytoImageList object to visualize the combined expression of up to six markers as composite images](imgs/Overview.png)

The `cytomapper` package contains three broad functionalities:

* Visualization of pixel-intensities as composites of up to six channels (`plotPixels`)

* Visualization of cell-specific features on segmentation masks (`plotCells`)

* Interactive gating of cells and visualization of gated cells on images (`cytomapperShiny`)

The follwing demonstration gives an overview on (1) data handling and processing,
(2) data visualization and (3) interactive analysis.

We use [imaging mass cytometry](https://www.nature.com/articles/nmeth.2869) data
to highlight the functionality of the `cytomapper` package. However, any imaging
technology is supported as long as the data can be read into R (memory
restrictions and file type restrictions.)

The raw data has been processed using the [ImcSegmentationPipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline).
In this tutorial, we assume that image segmentation has been performed beforehand.

# Reading in the data

The following section describes how to read in images (e.g. in `.tiff` format)
into `CytoImageList` objects and how to generate `SingleCellExperiment`
objects from these images.

## Reading in images

The `cytomapper::loadImages` function reads in multi-channel images and segmentation masks into `CytoImageList` objects.

```{r, reading-in-data-1, message=FALSE}
library(cytomapper)

# Read in 32-bit multi-channel images
(images <- loadImages("../data/images/", pattern = ".tiff"))

# Read in 16-bit unsigned integer segmentation masks
(masks <- loadImages("../data/masks/", pattern = ".tiff"))
```

It is always recommended to observe the numeric pixel values to make sure that
images were read in correctly:

```{r histograms}
# multi-channel images - first image, first channel
hist(log10(images[[1]][,,1] + 1))

# Segmentation mask
masks[[1]]
```

We notice, that the segmentation masks were not read in as integer images. 
This behavior arises due to issues with accessing the `tiff` metadata after
pre-processing using `CellProfiler`.

In these cases, the `cytomapper::scaleImages` function can be used to rescale
segmentation masks to only contain interger IDs. Here, the scaling factor
is `2 ^ 16 - 1 = 65535` accounting for the 16-bit unsigned integer encoding.

```{r scaleImages}
masks <- scaleImages(masks, 2 ^ 16 - 1)

masks[[1]]
```

As an alternative, while reading in the images, the `as.is` option can be set to `TRUE`.

```{r as.is}
masks <- loadImages("../data/masks/", pattern = ".tiff", as.is = TRUE)

masks[[1]]
```

We can already visualize the segmentation mask to get an idea of the tissue 
structure:

```{r mask-viz}
plotCells(masks)
```

### Reading data to disk

To increase scalability of the `cytomapper` package, images can be stored on
disk making use of the `HDF5Array` and `DelayedArray` package. Reading in the
data can also be parallelised, which is only recommended for large images.

```{r read-in-to-disk}
format(object.size(images), units = "Kb")

images_ondisk <- loadImages("../data/images/", pattern = ".tiff", 
                            on_disk = TRUE, h5FilesPath = "../data/images",
                            BPPARAM = BiocParallel::bpparam())

format(object.size(images_ondisk), units = "Kb")
```

All `cytomapper` functions support images and masks stored on disk.

## The `SingleCellExperiment` object

In the original analysis, single-cell features have been extracted using
`CellProfiler`. For this tutorial, we provide a `SingleCellExperiment` object,
which already contains the mean intensities per cell and channel and all
relevant metadata (e.g. cell-type annotation).

The `SingleCellExperiment` can be read-in form the `data` folder:

```{r read-in-sce}
(sce <- readRDS("../data/sce.rds"))

colData(sce)
```

# Image pre-processing

The following section will discuss setting image metadata and image normalization.

## Setting metadata

Before image visualization, there are a few metadata that need to be set.

We will need to set the channel names of the images via the `channelNames`
getter/setter function. The channel order here is the same as the row order of
the `SingleCellExperiment` object. We will also need to synchronise the image
IDs across the multi-channel images and segmentation masks by storing a
DataFrame in the elementMetadata slot of the CytoImageList object.

```{r, format-the-data, message=FALSE}
# Add channel names
channelNames(images) <- rownames(sce)

# Add image name to metadata
(mcols(images) <- mcols(masks) <- DataFrame(ImageName = c("E30", "G23", "J01")))
```

## Normalization

Channel normalization is a crucial step for image visualization to enhance
the visibility of biological signals. One option is to transform (e.g. `sqrt`)
transform the images. A more widely used alternative is to perform a range-scaling
and clipping of channel intensities.

The `cytomapper` package exports the `normalize` function that scales channels
between 0 and 1 across all images (by default). Clipping is performed by setting
the `inputRange` parameter in the `normalize` function.

```{r normalization}
# Min-max normalization
images_norm <- normalize(images)

# Clip the images
images_norm <- normalize(images_norm, inputRange = c(0, 0.2))
```

## Measuring object features

The `cytomapper::measureObjects` function takes the segmentation masks and
multi- channel image objects and  computes morphological (e.g. cell shape, size
and location) and intensity features (default: mean intensity per channel and
object/cell).

```{r, measure-features}
sce_measured <- measureObjects(mask = masks, image = images, img_id = "ImageName")
sce_measured
```

The pixel intensities per cell can be summarized in different ways (e.g. as
quantiles). Furthermore, parallelization is possible by setting `BPPARAM = bpparam()`. 
Cell-specific morphological features are stored in `colData(sce)` while the 
mean pixel intensities per cell and channel are stored in `counts(sce)`.

```{r show-internal-structure}
colData(sce_measured)

counts(sce_measured)[1:10, 1:10]
```

# Data visualization

After having formatted the data, we can now move on to data visualization.

## Visualizing pixel intensities

The `cytomapper::plotPixels` function is the main function for visualization
of pixel intensities.

The easiest way of visualizing images is to just specify channel names:

```{r plotPixels-1}
plotPixels(images, 
           colour_by = c("PIN", "CDH"))
```

Here, the intensities appear weak. For this, we have normalized the images:

```{r plotPixels-2}
plotPixels(images_norm, 
           colour_by = c("PIN", "CDH"))
```

We can see a decline in pro-insulin expression while E-cadherin expression stays the same.

Instead of normalizing the images, you can also specify the background (b), 
contrast (c) and gamma (g) level of the images:

```{r plotPixels-3}
plotPixels(images, 
           colour_by = c("PIN", "CDH"),
           bcg = list(PIN = c(0, 5, 1),
                      CDH = c(0, 5, 1)))
```

The `cytomapper` package allows flexible adjustments of all visual attributes of the images:

```{r plotPixels-4}
plotPixels(
    image = images,
    colour_by = c("PIN", "CD4", "CD8a"), 
    colour = list(PIN = c("black", "yellow"),
                  CD4 = c("black", "blue"),
                  CD8a = c("black", "red")),
    bcg = list(PIN = c(0, 10, 1),
               CD4 = c(0, 8, 1),
               CD8a = c(0, 10, 1)),
    image_title = list(
        text = c("Non-diabetic",
                 "Recent onset T1D",
                 "Long duration T1D")
        ),
    scale_bar = list(
        length = 100,
        label = expression("100 " ~ mu * "m")
        ))
```

The `colour_by` parameter defines the channel names by which to colour the
composite. Per channel, a colour scale is generated by setting `colour`. The
attributes of the image titles can be set via the parameter `image_title` and
attributes of the scale bar are set via the parameter `scale_bar`.

To see all available parameter options, refer to `?plotting-param`

## Visualizing cells

In the next step, we can visualize mean pixel intensities and cell-specific metadata
directly on the segmentation masks.

For this, we will use the `cytomapper::plotCells` function while combining the
`SingleCellExperiment` object and the `CytoImageList` storing the segmentation
masks:

```{r plotCells-1}
plotCells(mask = masks, object = sce, 
          cell_id = "CellNumber", img_id = "ImageName",
          colour_by = c("PIN", "CD8a", "CD4"), 
          exprs_values = "exprs")
```

The `assay(sce, "exprs")` slot stores the asinh-transformed mean pixel intensities 
per channel and cell. The `cell_id` entry needs to be stored in the `colData(sce)`
slot to link each individual cell to their corresponding pixels in the segmentation mask.
The `img_id` parameter specifies the `colData(sce)` entry storing the name/ID
of the image to which each cell belongs. This should also be an entry in 
`mcols(masks)`.

The `SingleCellExperiment` object is subsettable to highlight only certain cells:

```{r plotCells-2}
cur_sce <- sce[,sce$CellType %in% 
    c("beta", "alpha", "delta", "Tc", "Th")]

plotCells(mask = masks, object = cur_sce, 
          cell_id = "CellNumber", img_id = "ImageName",
          colour_by = c("PIN", "CD8a", "CD4"), 
          exprs_values = "exprs",
          missing_colour = "white")
```

Here, `missing_colour` defines the colour of cells missing in the `SingleCellExperiment` object.

This can also be useful when visualizing cell-specific metadata:

```{r plotCells-3}
plotCells(
    mask = masks,
    object = cur_sce,
    cell_id = "CellNumber",
    img_id = "ImageName",
    colour_by = "CellType",
    image_title = list(
        text = c("Non-diabetic",
                 "Recent onset T1D",
                 "Long duration T1D"),
        colour = "black"),
    scale_bar = list(
        length = 100,
        label = expression("100 " ~ mu * "m"),
        colour = "black"),
    missing_colour = "white",
    background_colour = "gray")
```

## Visualizing cells and pixel intensities

Finally, all three objects (multi-channel images, segmentation masks and
the single-cell data) can be combined to outline cells on composite images.

As a first demonstration, we select only pancreatic beta cells and visualize 
their corresponing marker pro-insulin:

```{r outline-1}
cur_sce <- sce[,sce$CellType == "beta"]

plotPixels(images_norm,
           mask = masks,
           object = cur_sce,
           cell_id = "CellNumber",
           img_id = "ImageName",
           colour_by = c("PIN", "H3"),
           bcg = list(H3 = c(0, 2, 1)),
           colour = list(PIN = c("black", "yellow"),
                         H3 = c("black", "blue"),
                         CellType = c(beta = "red")),
           outline_by = "CellType")
```

One can now also visualize multiple cell types and multiple markers together:

```{r outlineCells-2, echo=FALSE}
cur_sce <- sce[,sce$CellType %in% 
    c("beta", "alpha", "delta", "Tc", "Th")]

plotPixels(image = images,
           object = cur_sce,
           mask = masks,
           cell_id = "CellNumber",
           img_id = "ImageName",
           colour_by = c("PIN", "CD4", "CD8a"), 
           outline_by = "CellType",
           colour = list(PIN = c("black", "yellow"),
                         CD4 = c("black", "blue"),
                         CD8a = c("black", "red")),
           bcg = list(PIN = c(0, 10, 1),
                      CD4 = c(0, 8, 1),
                      CD8a = c(0, 10, 1)),
           image_title = list(text = c("Non-diabetic",
                                       "Recent onset T1D",
                                       "Long duration T1D")),
           scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m")),
           thick = TRUE)
```

# Interactive visualization

We have developed a `shiny` application to gate cells based on their mean
pixel intensity with additional visualization on the images.

The idea behind this gating strategy is to generate ground truth cell type labels, 
which can be used to train a classifier and classify cells rather than 
relying on clustering.

The shiny app can be opened using the `cytomapper::cytomapperShiny` function:

```{r cytomapperShiny}
if (interactive()) {
    
    cytomapperShiny(sce, mask = masks, image = images,
                    img_id = "ImageName", cell_id = "CellNumber")
    
}
```

**Using the Shiny application**

This help page provides a recommended workflow on how to most efficiently use the app. The workflow is solely a recommendation - the app provides full flexibility to change settings during each step. To see the full documentation, please refer to the help page found at `?cytomapperShiny`

1. Select the number of plots  
The slider under General controls can be used to specify the number of plots on which to perform gating. Up to two markers can be visualized per plot.

2. Select the sample  
The assay dropdown selection under General controls allows the user to specify on which assay entry to perform gating. In most cases, a log- or arcsinh-transformation can help to distinguish between 'positive' and 'negative' populations.

3. Select the markers  
For each plot, up to two markers can be specified. If selecting a single marker, please specify this marker in the first of the two dropdown menus. A violin plot is used to visualize the expression of a single marker while a scatter plot is used to visualize the expression of two markers.

4. Gate cells  
When selecting cells in one plot, only those cells are visualized on the following plot. Once markers, the assay or the number of plots are changed, gates are cleared.

5. Observe the selected cells  
After gating, the selected cells are visualized on the corresponding images by switching to the Images tab. By default, the first marker is selected. The user can change the displayed marker or press reset marker to switch to the markers used for gating. If a multi-channel image object is provided, the contrast of the image can be changed. The right panel visualizes the selected cells either by filling in the segmentation masks or by outlining the cells on the images.

6. Change samples  
Samples can now be iteratively changed using the dropdown menu under General controls . The gates will remain on the plots and can be adjusted for each sample.

7. Save the selected cells  
Finally, the selected cells can be saved by clicking the download button next to the '?' symbol. The selected cells will be stored as a `SingleCellExperiment` object in .rds format. Per selection, the user can provide a Cell label that will be stored in the `colData` under the `cytomapper_CellLabel` entry of the downloaded object.

# Further resources

For pre-processing multiplexed imaging data, please refer to the [ImcSegmentationPipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline) and/or the
[steinbock](https://github.com/BodenmillerGroup/steinbock) package.

To test the `cytomapper` package on different datasets, check out the [imcdatasets](https://bioconductor.org/packages/release/data/experiment/html/imcdatasets.html) Bioconductor package.

The [imcRtools](https://github.com/BodenmillerGroup/imcRtools) package is currently being developed
to facilitate handling of multiplexed imaging data and spatial analysis.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
