---
title: "EuroBioc2023 - cytomapper/cytoviewer: R/Bioconductor packages for visualization and exploration of highly multiplexed imaging data"
date: "`r BiocStyle::doc_date()`"
author:
- name: Lasse Meyer 
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: lasse.meyer@uzh.ch
- name: Nils Eling
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: nils.eling@uzh.ch
- name: Bernd Bodenmiller
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
output:
    BiocStyle::html_document:
        toc_float: yes
        pandoc_args: [
            "--output=index.html"
            ]
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(dirname(inputFile),'/index.html')) })
abstract: |
    Highly multiplexed imaging allows simultaneous spatially and single-cell resolved detection of dozens of biological molecules (e.g. proteins) in their native tissue context. As a result, these technologies allow an in-depth analysis of complex systems and diseases such as cancer. Here, we showcase two related R/Bioconductor packages, cytomapper and cytoviewer, that contain user-friendly functions to visualize and explore the multiplexed read-outs and cell-level information obtained by highly multiplexed imaging data. Firstly, the cytomapper package allows visualization of multi-channel pixel-level information as well as display of single cell-level information on segmentation masks. Secondly, the cytoviewer package builds on top of the cytomapper package and extends the static visualization strategies via an interactive Shiny application. The cytoviewer interface is divided into image-level and cell-level visualization. Users can overlay individual images with segmentation masks, visualize cell-specific metadata and download images. Both packages integrate well into the Bioconductor framework for single-cell and image analysis leveraging the image handling and analysis strategies from the EBImage Bioconductor package and building on commonly used classes such as SingleCellExperiment, SpatialExperiment and CytoImageList. Taken together, the Bioconductor packages cytomapper and cytoviewer provide a versatile and well-integrated toolbox for highly multiplexed imaging data visualization in R.
vignette: |
    %\VignetteIndexEntry{"cytomapper/cytoviewer: R/Bioconductor packages for visualization and exploration of highly multiplexed imaging data"}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: library.bib
---

`r fontawesome::fa(name = "github", fill = "#333")` <a href="https://github.com/lassedochreden">\@lassedochreden</a>  

`r fontawesome::fa(name = "github", fill = "#333")` <a href="https://github.com/nilseling">\@nilseling</a>  
`r fontawesome::fa(name = "twitter", fill = "#1DA1F2")` <a href="https://twitter.com/NilsEling">\@NilsEling</a> 

```{r setup, include=FALSE}
options(timeout=10000)
```

# Data and code availability

To follow this tutorial, please visit
[https://github.com/BodenmillerGroup/demos/tree/main/docs](https://github.com/BodenmillerGroup/demos/tree/main/docs).
The compiled `.html` file of this workshop is hosted at:
[https://bodenmillergroup.github.io/demos](https://bodenmillergroup.github.io/demos).

To reproduce the analysis, clone the repository:

```
git clone https://github.com/BodenmillerGroup/demos.git
```

and open the `EuroBioc2023_workshop.Rmd` file in the `docs` folder.

We will need to install the following packages for the workshop:

```{r installation, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("cytomapper", "cytoviewer")
```

```{r installation complete, eval=FALSE, message=FALSE}
library(cytomapper)
library(cytoviewer)
```

# Introduction

## Highly multiplexed imaging 

Highly multiplexed imaging allows simultaneous spatially and single-cell
resolved detection of dozens of biological molecules (e.g. proteins) in
their native tissue context. As a result, these technologies allow an
in-depth analysis of complex systems and diseases such as the tumor
microenvironment [@Jackson2020] and type 1 diabetes progression
[@Damond2019].

Imaging-based spatial proteomics methods [@Moffitt2022] can be broadly
divided into fluorescent cyclic approaches such as tissue-based cyclic
immunofluorescence (t-CyCIF) [@Lin2018] and one-step mass-tag based
approaches such as multiplexed ion beam imaging (MIBI) [@Angelo2014] and
IMC [@Giesen2014].

Across technologies, the acquired data are commonly stored as
multi-channel images, where each pixel encodes the abundance of all acquired
markers at a specific position in the tissue. Of note, the instructions below 
will focus on the visualization and exploration of **IMC data** as an example. 
However, data from other technologies such as t-CyCIF or MIBI, which produce 
pixel-level intensities and (optionally) segmentation masks, can be 
interactively visualized with `cytomapper` and `cytoviewer`.

### Imaging mass cytometry

IMC, an advancement of CyTOF, combines antibodies tagged with
isotopically pure rare earth metals with laser ablation and
mass-spectrometry-based detection to produce high-dimensional images
[@Giesen2014]. It captures the spatial expression of over 40 proteins in
parallel at a sub-cellular resolution of 1 μm. Thus, IMC is able to
detect cytoplasmic and nuclear localization of proteins.

## Highly multiplexed image analysis

When performing multiplexed image analysis, the user is
often faced with a diverse set of computational tools and complex analysis
scripts. The main analysis steps, irrespective of the biological question,
include 1) Visual inspection of images for quality control, 2) Image
pre-processing and segmentation and 3) Single-cell and spatial analysis.  

We developed an interoperabale, modularized computational end-to-end workflow 
[@Windhager2021] to process and analyze multiplexed imaging data **(Figure 1)**. 

The *steinbock* framework facilitates multi-channel image processing including 
raw data pre-processing, image segmentation and feature extraction. Data generated 
by *steinbock* can be directly read by the *imcRtools* R/Bioconductor package for 
data visualization and spatial analysis **(Figure 1)**. 

The *cytomapper* package [@Eling2020] supports image handling and composite as well as 
segmentation mask visualization, while the newly developed *cytoviewer* package 
[@Meyer2023] allows interactive and easy-to-use image visualization. Both packages 
integrate into the Bioconductor framework [@Gentleman2004] for single-cell and image analysis 
leveraging the image handling and analysis strategies from the *EBImage* Bioconductor 
package [@Pau2010] and building on commonly used Bioconductor classes such as 
*SingleCellExperiment* [@Amezquita2019], *SpatialExperiment* [@Righelli2022] and 
*CytoImageList* [@Eling2020]. 


![**Figure 1: Data output after image processing and its representation and visualization in R.**  
During image processing with the *steinbock* framework, all provided multi-channel 
images are first segmented to detect individual cells stored as segmentation masks. 
Based on the segmentation masks, single-cell features including the mean pixel 
intensity per cell and channel, morphological features and the x-y positions of 
the cells (regionprops), and spatial cell graphs are extracted. To work with the 
generated data in R, all multi-channel images are read into a single *CytoImageList* 
object, all segmentation masks are read into a single *CytoImageList* object and 
all single-cell features of all images are read into a single *SpatialExperiment* 
(or *SingleCellExperiment*) object. Patient and image level features are stored 
in the *elementMetadata* slot of each *CytoImageList* object and the *colData* 
slot of the *SpatialExperiment* object. For image visualization in R, 
*cytomapper* [@Eling2020] and *cytoviewer* [@Meyer2023] are used.](imgs/DataOutputR.png) 




## Workshop outline

In this workshop, we showcase two related R/Bioconductor packages, 
*cytomapper* [@Eling2020] and *cytoviewer* [@Meyer2023], that contain user-friendly 
functions to visualize and explore the multiplexed read-outs and cell-level information 
obtained by highly multiplexed imaging data. 

Outline: 

1. Download and exploration of example IMC data
2. cytomapper 
3. cytoviewer 
4. Future developments 

The visualization approaches presented here were taken from the [IMC data analysis
book](https://bodenmillergroup.github.io/IMCDataAnalysis/). The book provides much
more detailed information on most relevant IMC data analysis steps.



# Download and exploration of example data

## Download example dataset

Here, we will use an example IMC cancer dataset to showcase the functionality of 
`cytomapper` and `cytoviewer`. This
dataset was generated as part of the Integrated iMMUnoprofiling of large
adaptive CANcer patient cohort project ([immucan.eu](immucan.eu)) and
includes IMC data for 4 cancer patients diagnosed with different tumor
types (head and neck cancer, breast cancer, lung cancer and colorectal
cancer). 

The data input objects were processed as outlined in the [IMC data analysis
book](https://bodenmillergroup.github.io/IMCDataAnalysis/) and can
be downloaded from <https://zenodo.org/record/7647079>.

```{r download, message=FALSE}
# Download
download.file("https://zenodo.org/record/7647079/files/spe.rds",
              destfile = "spe.rds")
download.file("https://zenodo.org/record/7647079/files/images.rds",
              destfile = "images.rds")
download.file("https://zenodo.org/record/7647079/files/masks.rds",
              destfile = "masks.rds")
```

## Read data into R

### Images

The image data was stored as a `CytoImageList` object containing the
spillover corrected multi-channel images (n=14). Each image contains 40
channels and each channel represents the pixel-intensities of one marker
(proteins for IMC). The proteins for this dataset are immuno-oncology
related targets including Ecad, CD8a and CD68, which mark tumor, CD8+ T
cells and myeloid cells, respectively.

```{r images, message=FALSE, warning=FALSE}
# Load images
images <- readRDS("images.rds")
images
```

The `elementMetadata` slot of the `CytoImageList` object stores sample and 
patient ID as well as indication information. 

```{r metadata}
mcols(images)
```

The `channelNames` slot stores the names of the targeted proteins.

```{r channelNames}
channelNames(images)
```

The `CytoImageList` object inherits from a `SimpleList` object and therefore
supports common subsetting operations:

```{r subsetting}
images$Patient1_001
```

Here, individual images are stored in memory as `EBImage::Image` objects; 
however the `cytomapper` package also supports [storing images on disk](https://www.bioconductor.org/packages/release/bioc/vignettes/cytomapper/inst/doc/cytomapper_ondisk.html).

### Segmentation masks

The segmentation masks were again stored as a `CytoImageList` object containing
one mask (n=14) for each image. Segmentation masks are defined as one-channel
images containing integer values for cells and zero for background.

```{r masks}
# Load masks
masks <- readRDS("masks.rds")
masks
```

It is crucial that the image names are stored in the `elementMetadata` slot
for matching between multi-channel images and segmentation masks.

```{r check-metadata}
mcols(images)
mcols(masks)
```

### SpatialExperiment object

The single-cell information was stored in a `SpatialExperiment` object. Each
column entry represents a single-cell and each row entry represents a single
marker. We quantify protein abundance as the mean pixel intensity per marker and
cell.

```{r spe}
# Load spe
spe <- readRDS("spe.rds")
spe
```

It also contained various metadata information in the `colData` slot generated
during the analysis pipeline including patient-level information (such
as indication) and cell-level information (such as cell type and cell
area).

```{r colData}
# Explore colData
unique(spe$patient_id)
unique(spe$sample_id)
unique(spe$indication)
unique(spe$celltype)
range(spe$area)
```

# cytomapper 

The
[cytomapper](https://www.bioconductor.org/packages/release/bioc/html/cytomapper.html)
R/Bioconductor package was developed to support the handling and visualization
of multiple multi-channel images and segmentation masks [@Eling2020]. The main
data object for image handling is the
[CytoImageList](https://www.bioconductor.org/packages/release/bioc/vignettes/cytomapper/inst/doc/cytomapper.html#5_The_CytoImageList_object)
container to store multi-channel
images and segmentation masks.  

We will randomly select 3 images for visualization purposes.

```{r select-images}
# Load package 
library(cytomapper)

# Sample images
set.seed(220517)
cur_id <- sample(unique(spe$sample_id), 3)

cur_spe <- spe[,spe$sample_id %in% cur_id]
cur_images <- images[names(images) %in% cur_id]
cur_masks <- masks[names(masks) %in% cur_id]
```

## Pixel-level visualization 

The following section gives examples for visualizing individual channels or 
multiple channels as pseudo-color composite images. For this the `cytomapper`
package exports the `plotPixels` function which expects a `CytoImageList` object
storing one or multiple multi-channel images. In the simplest use case, a 
single channel can be visualized as follows:

```{r single-channel}
plotPixels(cur_images, 
           colour_by = "Ecad",
           bcg = list(Ecad = c(0, 5, 1)))
```

The plot above shows the tissue expression of the epithelial tumor marker
E-cadherin on the 3 selected images. The `bcg` parameter (default `c(0, 1, 1)`)
stands for "background", "contrast", "gamma" and controls these attributes of
the image. This parameter takes a named list where each entry specifies these
attributes per channel. The first value of the numeric vector will be added to
the pixel intensities (background); pixel intensities will be multiplied by the
second entry of the vector (contrast); pixel intensities will be exponentiated
by the third entry of the vector (gamma). In most cases, it is sufficient to
adjust the second (contrast) entry of the vector.

The following example highlights the visualization of 6 markers (maximum allowed
number of markers) at once per image. The markers indicate the spatial
distribution of tumor cells (E-cadherin), T cells (CD3), B cells (CD20), CD8+ T
cells (CD8a), plasma cells (CD38) and proliferating cells (Ki67).

```{r 6-channel}
plotPixels(cur_images, 
           colour_by = c("Ecad", "CD3", "CD20", "CD8a", "CD38", "Ki67"),
           bcg = list(Ecad = c(0, 5, 1),
                      CD3 = c(0, 5, 1),
                      CD20 = c(0, 5, 1),
                      CD8a = c(0, 5, 1),
                      CD38 = c(0, 8, 1),
                      Ki67 = c(0, 5, 1)))
```

### Adjusting colors

The default colors for visualization are chosen by the additive RGB (red, green,
blue) color model. For six markers the default colors are: red, green, blue,
cyan (green + blue), magenta (red + blue), yellow (green + red). These colors
are the easiest to distinguish by eye. However, you can select other colors for
each channel by setting the `colour` parameter:

```{r setting-colors}
plotPixels(cur_images, 
           colour_by = c("Ecad", "CD3", "CD20"),
           bcg = list(Ecad = c(0, 5, 1),
                      CD3 = c(0, 5, 1),
                      CD20 = c(0, 5, 1)),
           colour = list(Ecad = c("black", "burlywood1"),
                         CD3 = c("black", "cyan2"),
                         CD20 = c("black", "firebrick1")))
```

The `colour` parameter takes a named list in which each entry specifies the
colors from which a color gradient is constructed via `colorRampPalette`. These
are usually vectors of length 2 in which the first entry is `"black"` and the
second entry specifies the color of choice.

### Outlining cells on images 

The following section highlights the combined visualization of pixel- and
cell-level information at once. 

First, we can outline all cells on composite images to visually check 
segmnetation quality.

```{r outlining-1}
plotPixels(image = cur_images,
           mask = cur_masks,
           img_id = "sample_id",
           colour_by = "DNA1",
           bcg = list(DNA1 = c(0, 5, 1)),
           colour = list(DNA1 = c("black", "blue")),
           thick = TRUE)
```

In the next step, we can color the cell outlines by a metadata entry. By
specifying the `outline_by` parameter, the outlines of cells can now be colored
based on their metadata.

The following example first generates a 3-channel composite images displaying
the expression of E-cadherin, CD3 and CD20 before coloring the cells' outlines
by their cell phenotype.

```{r outlining-all-cells}
plotPixels(image = cur_images,
           mask = cur_masks,
           object = cur_spe, 
           cell_id = "ObjectNumber", 
           img_id = "sample_id",
           colour_by = c("Ecad", "CD3", "CD20"),
           outline_by = "celltype",
           bcg = list(Ecad = c(0, 5, 1),
                      CD3 = c(0, 5, 1),
                      CD20 = c(0, 5, 1)),
           colour = list(celltype = metadata(cur_spe)$color_vectors$celltype),
           thick = TRUE)
```

Distinguishing individual cell phenotypes is nearly impossible in the images
above.

However, the `SpatialExperiment` object can be subsetted to only contain cells
of a single or few phenotypes. This allows the selective visualization of cell
outlines on composite images.

Here, we select all CD8+ T cells from the dataset and outline them on a 2-channel
composite image displaying the expression of CD3 and CD8a. 

```{r outlining-CD8}
CD8 <- cur_spe[,cur_spe$celltype == "CD8"]

plotPixels(image = cur_images,
           mask = cur_masks,
           object = CD8, 
           cell_id = "ObjectNumber", img_id = "sample_id",
           colour_by = c("CD3", "CD8a"),
           outline_by = "celltype",
           bcg = list(CD3 = c(0, 5, 1),
                      CD8a = c(0, 5, 1)),
           colour = list(celltype = c("CD8" = "white")),
           thick = TRUE)
```

This type of visualization allows the quality control of two things: 1.
segmentation quality of individual cell types can be checked and 2. cell
phenotyping accuracy can be visually assessed against expected marker expression.

## Cell-level visualization 

In the following section, we will show examples on how to visualize single cells
as segmentation masks. This type of visualization allows to observe the spatial
distribution of cell phenotypes and the visual assessment of morphological
features.

### Visualizing metadata

The `cytomapper` package provides the `plotCells` function that accepts a
`CytoImageList` object containing segmentation masks. These are defined as
single channel images where sets of pixels with the same integer ID identify
individual cells. This integer ID can be found as an entry in the `colData(spe)`
slot and as pixel information in the segmentation masks. The entry in
`colData(spe)` needs to be specified via the `cell_id` argument to the
`plotCells` function. In that way, data contained in the `SpatialExperiment`
object can be mapped to segmentation masks. For the current dataset, the cell
IDs are stored in `colData(spe)$ObjectNumber`.

As cell IDs are only unique within a single image, `plotCells` also requires
the `img_id` argument. This argument specifies the `colData(spe)` as well as the
`mcols(masks)` entry that stores the unique image name from which each cell was
extracted. In the current dataset the unique image names are stored in
`colData(spe)$sample_id` and `mcols(masks)$sample_id`.

Providing these two entries that allow mapping between the `SpatialExperiment`
object and segmentation masks, we can now color individual cells based on their
cell type:

```{r celltype}
plotCells(cur_masks,
          object = cur_spe, 
          cell_id = "ObjectNumber", 
          img_id = "sample_id",
          colour_by = "celltype")
```

If only individual cell types should be visualized, the `SpatialExperiment`
object can be subsetted (e.g., to only contain CD8+ T cells). In the following
example CD8+ T cells are colored in red and all other cells that are not
contained in the dataset are colored in white (as set by the `missing_color`
argument).

```{r selective-visualization}
CD8 <- cur_spe[,cur_spe$celltype == "CD8"]

plotCells(cur_masks,
          object = CD8, 
          cell_id = "ObjectNumber", 
          img_id = "sample_id",
          colour_by = "celltype",
          colour = list(celltype = c(CD8 = "red")),
          missing_colour = "white")
```

In terms of visualizing metadata, any entry in the `colData(spe)` slot can be 
visualized. The `plotCells` function automatically detects if the entry 
is continuous or discrete. In this fashion, we can now visualize the area of each 
cell:

```{r area}
plotCells(cur_masks,
          object = cur_spe, 
          cell_id = "ObjectNumber", 
          img_id = "sample_id",
          colour_by = "area")
```



# cytoviewer

This section introduces the [cytoviewer](https://www.bioconductor.org/packages/release/bioc/html/cytoviewer.html)
R/Bioconductor package [@Meyer2023] for interactive
multi-channel image visualization. The `cytoviewer` package builds on 
top of the `r Biocpkg("cytomapper")` Bioconductor package and extends the 
static visualization strategies provided by `cytomapper` via an 
**interactive Shiny application**.

The graphical user interface of `cytoviewer` allows intuitive navigation and little 
coding experience is required to use the package. 

## Application overview

The `cytoviewer` interface is broadly divided into
**Image-level** (Composite and Channels) and
**Cell-level** (Masks) visualization (As seen above). It allows users to
overlay individual images with segmentation masks, integrates well with
`SingleCellExperiment` and `SpatialExperiment` objects for metadata
visualization and supports rapid `Image downloads` **(Figure 2B)**.

![**Figure 2: cytoviewer interface and functionality.**
**(A)** The supported functionality (right) of *cytoviewer* depends on the data 
inputs (left). To match information between the objects, cell (cell_id) and image 
(img_id) identifiers can be provided. SCE/SPE = *SingleCellExperiment*/*SpatialExperiment*. 
**(B)** The graphical user interface of *cytoviewer* is divided into a body, header, 
and sidebar. The body of *cytoviewer* includes the image viewer, which has three tabs: 
Composite (Image-level), Channels (Image-level), and Mask (Cell-level). Zooming is 
supported for Composite and Mask tabs. The package version, R session information, 
help page, and a drop-down menu for image downloads are located in the header. The 
sidebar menu has controls for sample selection, image visualization, mask visualization, 
and general settings. Scale bar: 150 µm **(C)** *cytoviewer* supports different viewing 
modes. Top: The “channels” tab of image-level visualization displays individual 
channels. Shown are Ecad (magenta), CD8a (cyan), and CD68 (yellow) marking tumor 
cells, CD8+ T cells, and myeloid cells, respectively. Center: The “composite” 
tab of image-level visualization visualizes image composites combining multiple 
channels. These composite images can be overlayed with cell outlines, which can 
be colored by cell-specific metadata. Shown here are cell outlines colored by 
cell area (continous value) and cell type (categorical value; tumor cells in 
white). Channel color settings are as follows for all markers: Contrast: 2,5; 
Brightness: 1; Gamma: 1.2. Bottom: The “mask” tab can be used to visualize 
segmentation masks that can be colored by cell-specific metadata. 
Shown here are segmentation masks colored by cell area (continuous) and cell 
type (categorical; tumor cells in magenta). Scale bars: 150 µm. **(D)** 
“Image appearance” controls can be used to add legends or titles and to change 
the scale bar length for image-level (top) and cell level (bottom) visualization. 
The cell-level mask plot depicts tumor (magenta), myeloid (yellow), and 
CD8+ T cells (cyan). Scale bars: 100 µm.
Adapted from [@Meyer2023]](imgs/cytoviewer.png)

### Data input format

The `cytoviewer` package combines objects of
`r Biocpkg("SingleCellExperiment")`, `r Biocpkg("SpatialExperiment")`
and `cytomapper::CytoImageList` classes (from `r Biocpkg("cytomapper")`)
to visualize image- and cell-level information.

The `cytoviewer` function call takes up to *five arguments* **(Figure 2A)**:

Firstly, `image` refers to a `CytoImageList` object containing one or
multiple multi-channel images where each channel represents the
pixel-intensities of one marker (proteins in IMC).

Secondly, `mask` refers to a `CytoImageList` object containing one or
multiple segmentation masks. Segmentation masks are defined as
one-channel images containing integer values, which represent the cell
ids or background.

Thirdly, the `object` entry refers to a `SingleCellExperiment` or
`SpatialExperiment` class object that contains cell-specific metadata in
the `colData` slot.

Lastly, to match information between the `CytoImageList` objects and the
`SingleCellExperiment`/`SpatialExperiment` object, two additional spots
can be specified:

-   `img_id`: a single character indicating the `colData` (of the
    `SingleCellExperiment`/`SpatialExperiment` object) and
    `elementMetadata` (of the `CytoImageList` object) entry that
    contains the image identifiers. These image ids have to match
    between the `SingleCellExperiment`/ `SpatialExperiment` object and
    the `CytoImageList` objects.

-   `cell_id`: a single character indicating the `colData` entry that
    contains the cell identifiers. These should be integer values
    corresponding to pixel-values in the segmentation masks.

### Data input variations

The functionality of `cytoviewer` depends on which input objects are
user-provided. Below we describe the **four use cases** in respect to
input objects and functionality **(Figure 2A)**.  

**1. Usage of cytoviewer with images, masks and object**

The *full* functionality of cytoviewer can be leveraged when `image`,
`mask` and `object` are provided, which is the main intended use case.

This allows image-level visualization (Composite and
Channels), cell-level visualization, overlaying images
with segmentation masks as well as metadata visualization.  

**2. Usage of cytoviewer with images only**

If only the `image` object is specified, image-level
visualization (Composite and Channels) is possible.  

**3. Usage of cytoviewer with images and masks**

Image-level visualization (Composite and Channels),
overlaying of images with segmentation masks and
Cell-level visualization is feasible when `image` and
`mask` objects are provided.  

**4. Usage of cytoviewer with masks and object**

If `mask` and `object` are specified, cell-level
visualization as well as metadata visualization is possible.

## Function call

Here as an example, we will call `cytoviewer` with the `image`, `mask` and
`object` data from the `cytomapper` section to leverage all provided functionality.

This setting allows image-level visualization (Composite
and Channels), cell-level visualization, overlaying images
with segmentation masks as well as metadata visualization.

For further details, please refer to the `?cytoviewer` manual or the
`Help page` within the shiny application.

```{r cytoviewer, message=FALSE, error=FALSE}
library(cytoviewer)

# Use cytoviewer with images, masks and object
app <- cytoviewer(image = cur_images, 
                  mask = cur_masks, 
                  object = cur_spe, 
                  img_id = "sample_id", 
                  cell_id = "ObjectNumber")

if (interactive()) {
  
  shiny::runApp(app)

  }
```

## Application overview 

## Image-level visualization

Image visualization control is split into *basic* and *advanced controls* 
**(Figure 2C)**.

*Basic controls* supports the selection of up to six markers/channels
for `image` display. Each marker has color control settings that allow
the user to set contrast, brightness, gamma and select a channel color.

In the *advanced controls* part, the user can choose to overlay the
displayed images with provided segmentation `masks`. Outline color and
mask thickness can be adjusted by the user. Moreover, the masks can be
outlined by cell-specific metadata provided in `colData` slot of the
`object`.

Of note, for categorical and continuous metadata entries the user can
choose between discrete colors and continuous color palettes (viridis,
inferno, plasma), respectively.

## Cell-level visualization

Cell visualization has *basic controls* **(Figure 2C)**.

Here, the user can choose to display the provided segmentation `masks`.
If an `object` is provided, the masks can be colored by cell-specific
metadata.

Please note again that for categorical and continuous metadata entries
the user can choose between discrete colors and continuous color
palettes (viridis, inferno, plasma), respectively.

## General controls

General controls is subdivided into an *Image appearance* and 
*Image filters* part **(Figure 2D)**.

In the *Image appearance* section, the user can adjust the scale bar
length and include legend/image titles, while the *Image filters*
section allows to control pixel-wise interpolation (default) and apply a
Gaussian filter.

## Image download

The `cytoviewer` package supports fast and uncomplicated image
downloads. Download controls are part of the **Header**. 

The user can specify a file name, select the image of interest
(Composite, Channels, Mask) and the file format (pdf, png). Upon
clicking the download button, a pop-window should appear where the user
can specify the download location.



# Future developments 

1. Future developments of both packages include an integration with modern imaging 
file types such as `OME-NGFF` and `spatialData` to enable image and single-cell 
analysis in a programming language agnostic fashion.  

2. We further envision to integrate interactive visualization of spatial cell graphs 
into `cytoviewer` ("Points-level").


# Further resources

The [IMC data analysis book](https://bodenmillergroup.github.io/IMCDataAnalysis/)
contains a detailed overview on the presented and other approaches for 
multiplexed image analysis and visualization.

The [steinbock](https://github.com/BodenmillerGroup/steinbock) framework 
provides functionality for image processing.

The [ImcSegmentationPipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline)
provides a GUI-based version of the segmentation pipeline based on Ilastik
pixel classification and image segmentation via CellProfiler.

The
[imcRtools](https://www.bioconductor.org/packages/release/bioc/html/imcRtools.html)
package supports reading single-cell data from segmented images, multi-channel 
spillover correction, and spatial data analysis.

The [imcdatasets](https://bioconductor.org/packages/release/data/experiment/html/imcdatasets.html) R/Bioconductor package contains a number of publically available IMC datasets.



# Acknowledgments

Nils Eling, Nicolas Damond and Tobias Hoch developed the `cytomapper` package. 
Lasse Meyer and Nils Eling developed the `cytoviewer` package. 

We thank Prof. Bernd Bodenmiller for his continued support. 

Nils Eling is funded by Marie Sklodowska Curie Actions.



# Session Info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```



# References 