---
title: "EuroBioc2023 - cytomapper/cytoviewer: R/Bioconductor packages for visualization and exploration of highly multiplexed imaging data"
date: "`r BiocStyle::doc_date()`"
author:
- name: Lasse Meyer 
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: lasse.meyer@dqbm.uzh.ch
- name: Nils Eling
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
- name: Bernd Bodenmiller
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
output:
    BiocStyle::html_document:
        toc_float: yes
        pandoc_args: [
            "--output=index.html"
            ]
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(dirname(inputFile),'/index.html')) })
abstract: |
    Highly multiplexed imaging allows simultaneous spatially and single-cell resolved detection of dozens of biological molecules (e.g. proteins) in their native tissue context. As a result, these technologies allow an in-depth analysis of complex systems and diseases such as cancer. Here, we showcase two related R/Bioconductor packages, cytomapper and cytoviewer, that contain user-friendly functions to visualize and explore the multiplexed read-outs and cell-level information obtained by highly multiplexed imaging data. Firstly, the cytomapper package allows visualization of multi-channel pixel-level information as well as display of single cell-level information on segmentation masks. In addition, it includes a Shiny application to enable hierarchical gating and visualization of selected cells. Secondly, the cytoviewer package builds on top of the cytomapper package and extends the static visualization strategies via an interactive Shiny application. The cytoviewer interface is divided into image-level and cell-level visualization. Users can overlay individual images with segmentation masks, visualize cell-specific metadata and download images. Both packages integrate well into the Bioconductor framework for single-cell and image analysis leveraging the image handling and analysis strategies from the EBImage Bioconductor package and building on commonly used classes such as SingleCellExperiment, SpatialExperiment and CytoImageList. Taken together, the Bioconductor packages cytomapper and cytoviewer provide a versatile and well-integrated toolbox for highly multiplexed imaging data visualization in R.
vignette: |
    %\VignetteIndexEntry{"cytomapper/cytoviewer: R/Bioconductor packages for visualization and exploration of highly multiplexed imaging data"}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: library.bib
---

`r fontawesome::fa(name = "github", fill = "#333")` <a href="https://github.com/lassedochreden">\@lassedochreden</a>  

`r fontawesome::fa(name = "github", fill = "#333")` <a href="https://github.com/nilseling">\@nilseling</a>  
`r fontawesome::fa(name = "twitter", fill = "#1DA1F2")` <a href="https://twitter.com/NilsEling">\@NilsEling</a> 

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Desktop/Github/demos/docs/")
options(timeout=10000)
```

# Data and code availability

To follow this tutorial, please visit
[https://github.com/BodenmillerGroup/demos/tree/main/docs](https://github.com/BodenmillerGroup/demos/tree/main/docs).
The compiled `.html` file of this workshop is hosted at:
[https://bodenmillergroup.github.io/demos](https://bodenmillergroup.github.io/demos).

We will need to install the following packages for the workshop:

```{r installation, eval=FALSE}
# 1. cytomapper 
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("cytomapper")

# 2. cytoviewer
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

remotes::install_github("BodenmillerGroup/cytoviewer")
```

```{r installation complete, eval=FALSE}
library(cytomapper)
library(cytoviewer)
```

To reproduce the analysis, clone the repository:

```
git clone https://github.com/BodenmillerGroup/demos.git
```

and open the `EuroBioc2023_workshop.Rmd` file in the `docs` folder.


# Introduction

## Highly multiplexed imaging 

Highly multiplexed imaging allows simultaneous spatially and single-cell
resolved detection of dozens of biological molecules (e.g. proteins) in
their native tissue context. As a result, these technologies allow an
in-depth analysis of complex systems and diseases such as the tumor
microenvironment [@Jackson2020] and type 1 diabetes progression
[@Damond2019].

Imaging-based spatial proteomics methods [@Moffitt2022] can be broadly
divided into fluorescent cyclic approaches such as tissue-based cyclic
immunofluorescence (t-CyCIF) [@Lin2018] and one-step mass-tag based
approaches such as multiplexed ion beam imaging (MIBI) [@Angelo2014] and
IMC [@Giesen2014].

Across technologies, the acquired data are commonly stored as
multi-channel images, where each pixel encodes the abundance of all acquired
markers at a specific position in the tissue. Of note, the instructions below 
will focus on the visualization and exploration of **IMC data** as an example. 
However, data from other technologies such as t-CyCIF or MIBI, which produce 
pixel-level intensities and (optionally) segmentation masks, can be 
interactively visualized with `cytomapper` and `cytoviewer`.

### Imaging mass cytometry

IMC, an advancement of CyTOF, combines antibodies tagged with
isotopically pure rare earth metals with laser ablation and
mass-spectrometry-based detection to produce high-dimensional images
[@Giesen2014]. It captures the spatial expression of over 40 proteins in
parallel at a sub-cellular resolution of 1 μm. Thus, IMC is able to
detect cytoplasmic and nuclear localization of proteins.

## Highly multiplexed image analysis

When performing end-to-end multiplexed image analysis, the user is
often faced with a diverse set of computational tools and complex analysis
scripts. The main analysis steps, irrespective of the biological question,
include 1) Visual inspection of images for quality control, 2) Image
pre-processing and segmentation and 3) Single-cell and spatial analysis.  

We developed an interoperabale, modularized computational end-to-end workflow 
[@Windhager2021] to process and analyze multiplexed imaging data **(Figure 1)**. 

The *steinbock* framework facilitates multi-channel image processing including 
raw data pre-processing, image segmentation and feature extraction. Data generated 
by *steinbock* can be directly read by the *imcRtools* R/Bioconductor package for 
data visualization and spatial analysis **(Figure 1)**. 

The *cytomapper* package [@Eling2020] support image handling and composite as well as 
segmentation mask visualization, while the newly developed *cytoviewer* package 
[@Meyer2023] supports interactive and easy-to-use image visualization. 

![**Figure 1: Overview of the multiplexed image processing and analysis workflow.** 
Raw image data can be interactively visualized using *napari* plugins such as
*napari-imc* for IMC, to assess data quality and for exploratory visualization.
The *steinbock* framework performs image pre-processing, cell segmentation and
single-cell data extraction using established approaches and standardized file
formats. Data can be imported into R using the *imcRtools* package, which
further supports spatial visualization and analysis. Storing the data in a
*SingleCellExperiment* or *SpatialExperiment* object, *imcRtools* integrates
with a variety of data analysis tools of the Bioconductor project such as
*cytomapper* [@Eling2020] and *cytoviewer* [@Meyer2023]. Alternatively, 
*steinbock* exports data to the *anndata* format for analysis in Python, e.g. 
using *squidpy*.](imgs/Overview.png)


## Workshop outline

In this workshop, we showcase two related R/Bioconductor packages, 
*cytomapper* [@Eling2020] and *cytoviewer* [@Meyer2023], that contain user-friendly 
functions to visualize and explore the multiplexed read-outs and cell-level information 
obtained by highly multiplexed imaging data. 

Outline: 

1. Download and exploration of example data
2. cytomapper 
3. cytoviewer 
4. Future developments 

The visualization approaches presented here were taken from the [IMC data analysis
book](https://bodenmillergroup.github.io/IMCDataAnalysis/). The book provides much
more detailed information most relevant IMC data analysis steps.


# Download and exploration of example data

## Download example dataset

Here, we will use an example IMC cancer dataset to showcase the functionality of 
`cytomapper` and `cytoviewer`. This
dataset was generated as part of the Integrated iMMUnoprofiling of large
adaptive CANcer patient cohort project ([immucan.eu](immucan.eu)) and
includes IMC data for 4 cancer patients diagnosed with different tumor
types (head and neck cancer, breast cancer, lung cancer and colorectal
cancer). 

The data input objects were processed with the [IMC data analysis
workflow](https://bodenmillergroup.github.io/IMCDataAnalysis/) and can
be downloaded from <https://zenodo.org/record/7079294>.

```{r download}
# Download
download.file("https://zenodo.org/record/7647079/files/spe.rds",
              destfile = "spe.rds")
download.file("https://zenodo.org/record/7647079/files/images.rds",
              destfile = "images.rds")
download.file("https://zenodo.org/record/7647079/files/masks.rds",
              destfile = "masks.rds")
```

## Read data into R

### Images

The image data was stored as a `CytoImageList` object containing the
spillover corrected multi-channel images (n=14). Each image contains 40
channels and each channel represents the pixel-intensities of one marker
(proteins for IMC). The proteins for this dataset are immuno-oncology
related targets including Ecad, CD8a and CD68, which mark tumor, CD8+ T
cells and myeloid cells, respectively.

```{r images}
# Load images
images <- readRDS("images.rds")
images
```

### Segmentation masks

The segmentation masks were again stored as a `CytoImageList` object containing
one mask (n=14) for each image. Segmentation masks are defined as one-channel
images containing integer values for cells and zero for background.

```{r masks}
# Load masks
masks <- readRDS("masks.rds")
masks
```

### Metadata object

The metadata object was stored in `SpatialExperiment` format. Each column entry 
represents a single-cell and each row entry represents a single marker. 
We quantify protein abundance as the mean pixel intensity per marker and cell. 
It also contained various metadata information in the `colData` slot generated
during the analysis pipeline including patient-level information (such
as indication) and cell-level information (such as cell type and cell
area).

```{r spe}
# Load spe
spe <- readRDS("spe.rds")
spe
```

```{r colData}
unique(spe$patient_id)
unique(spe$sample_id)
unique(spe$indication)
unique(spe$celltype)

counts(spe)[1:5,1:5]
assay(spe, "exprs")[1:5,1:5]
```


# cytomapper 

## Image-level visualization 

## Cell-level visualization 

## Adjusting plot annotations 

## Further functionality

# cytoviewer 

## Data input variations 

## Function call 

## Application overview 

# Future developments 

# Further resources

The [IMC data analysis book](https://bodenmillergroup.github.io/IMCDataAnalysis/)
contains a detailed overview on the presented and other approaches for 
multiplexed image analysis and visualization.

The [steinbock](https://github.com/BodenmillerGroup/steinbock) framework 
provides functionality for image processing.

The [ImcSegmentationPipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline)
provides a GUI-based version of the segmentation pipeline based on Ilastik
pixel classification and image segmentation via CellProfiler.

The
[imcRtools](https://www.bioconductor.org/packages/release/bioc/html/imcRtools.html)
package supports reading single-cell data from segmented images, multi-channel 
spillover correction, and spatial data analysis.

The [imcdatasets](https://bioconductor.org/packages/release/data/experiment/html/imcdatasets.html) R/Bioconductor package contains a number of publically available IMC datasets.


# Acknowledgments

Nils Eling, Nicolas Damond and Tobias Hoch developed the `cytomapper` package. 
Lasse Meyer and Nils Eling developed the `cytoviewer` package. 

We thank Prof. Bernd Bodenmiller for his continued support. 

Nils Eling is funded by Marie Sklodowska Curie Actions.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References 